---
title: "ggplot2_training"
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)

knitr::opts_chunk$set(echo = TRUE, message = TRUE)
```

* ggplot is the data visualization package
* based on *grammar of graphics*

* data via geom -> coordinate system -> plot...

```{r}
mpg
```

Shortcut for inserting code chunk in Rmd: alt-command-I
Can learn more about datasets in R by typing "?datasetname"

Arguements for ggplot: data, aesthetics (what to plot on axes); need to use a geom_ to tell ggplot *how* to plot

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) + geom_point()
```

```{r}
#assign this plot to a variable

car_plot <- ggplot(data = mpg, aes(displ, hwy))
```

Can use this approach to use same set up to do multiple plots

```{r}
car_plot + geom_point()
```

Remember, the "+" sign must be at the end of the line, cannot be at beginning of new indented line

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) +
  geom_point(alpha = 0.4)
```

Alpha adds transparency to the plotted data; this allows us to see that there are actually points stacked on top of each other here!

Another thing you can do is add color:

```{r}
ggplot(data = mpg, aes(x = displ, y = hwy)) + 
  geom_point(aes(color = class))
```

The "aes(color = class)" tells ggplot we want the points colored by the "class" of car

ggplot has a limited number of default shapes

```{r}
ggplot(data = mpg) + 
  geom_point(aes(x = displ, y = hwy, shape = class))
```

Specifying x and y in aesthetics in the ggplot() portion essentially names them as global aesthetics, and unless specify x and y in geom_ aes() arguements, will use those global specifications

```{r}
ggplot(data = mpg, aes(x = hwy, y = cty)) +
  geom_point(aes(size = class, color = fl))
```


