---
title: "Tidyr & Dplyr Training"
author: "Ellie Campbell"
date: "3/6/2018"
output: html_document
---


### Data Wrangling with dplyr


Data scientist spend too much time, about 50-80% of time doing collection and prep of data before useful exploration! 

Using tidyverse in R --> manipulating columns and rows

Wrangling is part of the "transform" or "tidy" phase of data analysis workflow

What is the difference between tidying and transforming?

### What is tidy data?

* Each variable in its own column
* Each observation saved in its own row

### The Tidyverse

A suite of R packages for data wrangling



***

# Gapminder Wrangle

We're going to wrangle gapminder data using `dplyr`.

```{r}
library(tidyverse) # install.packages('tidyverse') if get error saying package not found when this is run
```

When run "library(tidyverse)" get warning messages about functions masked by tidyverse

Will be using gapminder data from OHI-Science/data-science-training
Want raw csv data, can click "raw" tab and copy URL (https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv)


```{r}
# read in dataset from url:

gapminder <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')

# returns message about how data is parsed, i.e. was each column read as a string or as an integer, or as a double
```


## Explore Dataframe as a Whole:
```{r}
head(gapminder) # shows first 6 rows by default, but can specify more rows by including an additional parameter
tail(gapminder, 15)

str(gapminder) # structure of gapminder, gives information about dataframe; helpful to see what class each variable is so can determine whether can use different types of functions
names(gapminder) # tells us the column (variable) names

dim(gapminder) # tells the dimensions of the dataframe: number rows by number of columns
nrow(gapminder) # number of rows
ncol(gapminder) # number of columns

```


Most functions adn packages will expect your data to be in a dataframe; usually by default, data is read into a dataframe

Dataframes are awesome because they can contain both characters and numbers

Tibbles and tibble dataframes (`tbl` or `tbl_df`) are dataframes which can also contain other dataframes or things like ggplot

The function `c()` combines things into a list or a vector -- "combine" function

Tidyverse tries to keep everything consistent; if one class goes in, returns an object of the same class


## Summary Statistics for Dataframe:
```{r}
summary(gapminder) # for numeric values will return min median mean max and quartiles; for character vars will tell you it is a character class an the length

library(skimr) # run install.package('skimr') if have error 'package not found'
skim(gapminder) # cool package to look at data 'at a glance' and plays nicely with other tidyverse packages


```

## Explore Inside the Dataframe

```{r}
# gapminder$lifeExp gives information just for the single column in the dataframe, called "lifeExp" -- returns the entire column...
head(gapminder$lifeExp) # output is shown in integer vector
str(gapminder$lifeExp) # not esspecially enlightening looking at structure of just one column...
```


## Dplyr Basics:

* `filter()` subsets data by rows
* `select()` selects columns (i.e. variables)
* `mutate()` adds a variable based on other columns
* `summarize()` collapses many columns' values into a single new columnn

Will always return a dataframe

Work similarly, programatically

Can be strung together







