---
title: "Tidyr & Dplyr Training"
author: "Ellie Campbell"
date: "3/6/2018"
output: html_document
---


### Data Wrangling with dplyr


Data scientist spend too much time, about 50-80% of time doing collection and prep of data before useful exploration! 

Using tidyverse in R --> manipulating columns and rows

Wrangling is part of the "transform" or "tidy" phase of data analysis workflow

What is the difference between tidying and transforming?

### What is tidy data?

* Each variable in its own column
* Each observation saved in its own row

### The Tidyverse

A suite of R packages for data wrangling



***

# Gapminder Wrangle

We're going to wrangle gapminder data using `dplyr`.

```{r}
library(tidyverse) # install.packages('tidyverse') if get error saying package not found when this is run
```

When run "library(tidyverse)" get warning messages about functions masked by tidyverse

Will be using gapminder data from OHI-Science/data-science-training
Want raw csv data, can click "raw" tab and copy URL (https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv)


```{r}
# read in dataset from url:

gapminder <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')

# returns message about how data is parsed, i.e. was each column read as a string or as an integer, or as a double
```


## Explore Dataframe as a Whole:
```{r}
head(gapminder) # shows first 6 rows by default, but can specify more rows by including an additional parameter
tail(gapminder, 15)

str(gapminder) # structure of gapminder, gives information about dataframe; helpful to see what class each variable is so can determine whether can use different types of functions
names(gapminder) # tells us the column (variable) names

dim(gapminder) # tells the dimensions of the dataframe: number rows by number of columns
nrow(gapminder) # number of rows
ncol(gapminder) # number of columns

```


Most functions adn packages will expect your data to be in a dataframe; usually by default, data is read into a dataframe

Dataframes are awesome because they can contain both characters and numbers

Tibbles and tibble dataframes (`tbl` or `tbl_df`) are dataframes which can also contain other dataframes or things like ggplot

The function `c()` combines things into a list or a vector -- "combine" function

Tidyverse tries to keep everything consistent; if one class goes in, returns an object of the same class


## Summary Statistics for Dataframe:
```{r}
summary(gapminder) # for numeric values will return min median mean max and quartiles; for character vars will tell you it is a character class an the length

library(skimr) # run install.package('skimr') if have error 'package not found'
skim(gapminder) # cool package to look at data 'at a glance' and plays nicely with other tidyverse packages


```

## Explore Inside the Dataframe

```{r}
# gapminder$lifeExp gives information just for the single column in the dataframe, called "lifeExp" -- returns the entire column...
head(gapminder$lifeExp) # output is shown in integer vector
str(gapminder$lifeExp) # not esspecially enlightening looking at structure of just one column...
```


## Dplyr Basics:

* `filter()` subsets data by rows
* `select()` selects columns (i.e. variables)
* `mutate()` adds a variable based on other columns
* `summarize()` collapses many columns' values into a single new columnn

Will always return a dataframe

Work similarly, programatically

Can be strung together

```{r}

# Filter:

filter(gapminder, lifeExp <29) # want to filter to look at where life expectancy is less than 29 years old; filter based on condition specified in second arguement of the function
filter(gapminder, country == "Mexico")
filter(gapminder, country %in% c("Mexico", "Chile"))
filter(gapminder, country == "Mexico" | country == "Chile") # does the same thing as above

# Select:

select(gapminder, year, lifeExp) # select just the year and life expectancy columns
select(gapminder, -continent, -lifeExp) # can use minus to deselect certain columns

# Combining Select and Filter

gap_cambodia <- filter(gapminder, country == "Cambodia") # filter the country from Cambodia 
gap_cambodia_2 <- select(gap_cambodia, -continent, -lifeExp) # and remove continent and life expectancy columns..
gap_cambodia_2 # view results

```

## Pipe Operator!

**Life-changing** pipe operator: %>%, the keyboard shortcut is shift-command-n

Essentially like saying "and then"

This allows you to string together a bunch of functions, and have fewer intermediate variables

Tell it up front what data you will be using; that is carried throughout the next specified operations strung together with pipe operators

```{r}
gapminder %>% head() # equivalent to head(gapminder)
gapminder %>% head(3) # take gapminder data, and then return the first three rows; no longer need to specify the datam, because that is announced up front

gap_cambodia <- gapminder %>% filter(country == "Cambodia") %>% select(-continent, -lifeExp) # this is same as what we did above, but is cleaner since data is announced upfront and then operations are chained
gap_cambodia
```

## More Tidyverse Verbs

Add variables uing `mutate`:
```{r}
gapminder %>% 
  mutate(index = 1:nrow(gapminder)) # mutate takes dataframe and adds another column to it; index is name of new column and 1:nrow(gapminder) tells it to make rowvalues the vector 1 though the number of rows

gapminder %>% 
  mutate(gdp = pop * gdpPercap) # since already specified using gapminder dataset, don't have to use the $ notation to get individual dataframe column; also Rstudio will recognize now+ try to autofill colnames

```

Find the maximum gdpPercap of Egypt and Vietname, and save it in a new colunm:
```{r}
gapminder %>%
  filter(country %in% c("Egypt", "Vietnam")) %>% 
  mutate(max_gdp = max(gdpPercap)) # calculates the maximum of both Egypt and Vietnam TOGETHER...
  
```

To make max_gdp for each Egype and Vietnam (within each country respectively)...

### `group_by()`

```{r}
x <- gapminder %>%
  filter(country %in% c("Egypt", "Vietnam")) %>% # could put group_by before this filter statement, but then would force R to do more work, grouping more variables (more computationally intensive)
  group_by(country) %>%  # as long as group_by before mutating, will do mutate calculations on specified groups
  mutate(max_gdp = max(gdpPercap)) %>% # ALWAYS END A GROUP_BY WITH UNGROUP, OTHER WISE WILL CONSIDER CLUSTERS IN LATER STEPS
  ungroup()

str(x)
```


### `summarize()`

Helpful to use with `group_by`

```{r}
gapminder %>% 
  group_by(country) %>% 
  mutate(gdp = pop*gdpPercap) %>% 
  summarize(max_gdp = max(gdp)) %>% 
  ungroup() %>% 
  arrange(max_gdp) # rearranged order of rows, defaults to assending order (minimum to maximum) based on maximum gdp; can make into descending order by including "desc()" e.g. "arrange(desc(max_gdp)))"
```


Let's now find the maximum life expectancy for countries in Asia, hint: max()

```{r}
gapminder %>%
  filter(continent == "Asia") %>% 
  group_by(country) %>% 
  summarize(max_lifeExp = max(lifeExp)) %>% 
  ungroup()
```


# Joining Datasets

Combining datasets into one dataframe so can streamline analysis: Relational data/Joining data

Might be different ways in which you want to combine data

* Non-matching colunms (like arcgis join)
* Non-matching observations (like arcgis merge)

Ask yourself: what do you want to keep?

If want to merge additional observations with some differences in included variables, may end up with NAs

* Left-Join: join matching rows from B to A, keeping all observations from table A
* Right-Join: joing mtatching rows from A to B, keeping all observations from table B
* Inner-Join: Retain only rows in both sets; might result in some lost data
* Full-Join: Retain all values, all rows; can later filter out all rows with NAs in second step after viewing NAs


```{r}
co2 <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv') # read in CO2 data
co2 %>% head() # shows in rmd the first 6 rows of the co2 data

gap_2007 <- gapminder %>% 
  filter(year == 2007)

lj <- left_join(gap_2007, co2, by = "country") # join the co2 data to the gap2007 subset by the country variable, i.e. match by looking at the country names
lj %>% summary() # can see from the summary that there are 135 NAs out of 142 observations in the left join dataframe


rj <- right_join(gap_2007, co2, by = "country")
head(rj)


```



## Other random notes from training session...

Shortcut: control-alt-m opens commit window for github

If forget pipe operator and don't see right away can reindent lines ("Code" > "Reindent Lines"), shortcut: command I

Command-arrow to jump to the end of the code line


# `tidyr`

```{r}
library(tidyr)
gapminder <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')
gap_wide <- read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')
head(gapminder, 10)
head(gap_wide,10)
```

In a tidy dataset: each variable has its own column, and each observation has its own row

## Two Dataset Formats

Wide format: have multiple observation variables with the same type of data, e.g. with repeated observations over time or multiple columns for categorical variables of same type

Long format: one column for observed variable and others for ID variables

For humans, wide format is often more intuitive, but long format is often more "machine readable." Mathematical functions tend to be better on long format but wide format is sometimes better for plotting.

## Verbs for Tidying data with `tidyr`

* `gather()` -- turns columns into rows, i.e. from long to wide, similar to melt in the `reshape` package
* `spread()` -- turns rows to columns, i.e. from wide to long, similar to cast functions in `reshape`
* `separate()` -- 
* `unite()` -- 


This gapminder dataset is somewhere between long and wide data formats, with 3 ID variables (continent, year, country) and 3 observation variables (pop, lifeExp, gdpPercap)

Explore: str(gap_wide) and str(gapminder), or View(gap_wide)

Now: we wnat to turn the gap_wide dataset into the gapminder dataset

For long format, we want just one "key" variable that identifies the variable name and another "value" column for the data. For our gapminder data we'd like an *intermediate* format with gdpPercap, lifeExp, and pop as value columns. And we want a single column that specifies year, rather than 12x3=36 columns for each gdpPercap, lifeExp, and pop in each of the 12 years.


### `gather()`

Two ways to reshape into our desired intermediate format:
```{r}
gap_long <- gap_wide %>% 
  gather(key = obstype_year, value = obs_values,
         starts_with('pop'), starts_with('lifeExp'), starts_with('gdpPercap')) # these tell gather we want these variables left untouched; reshape around these
# could list column names instead of the starts_with arguments, but there are too many in this case

gap_long <- gap_wide %>% 
  gather(key = obstype_year, value = obs_values,
         -continent, -country) # here, you could instead of listing columns to include, exclude columns you don't want included in reshaping
```

This is still isn't quite what we wanted though... the obstype_year contains all the values for each gdpPercap, pop, and lifeExp, concatenated with the year of the observation, adjacent to the obs_values column which contains those actual values. To split this long format table so it has each of gdpPercap, pop, and lifeExp as a separate column containing its respective obs_value value, we need to use `separate()`

### `separate()`
```{r}
gap_intermed <- gap_long %>% 
  separate(obstype_year, # variable to separate
           into = c('obs_type', 'year'), # what to separate into
           sep = "_", # the character that separates the two pieces of information in obstype_year values
           convert = TRUE) # ensures that the year column is an integer rather than a string
```

We now have a year column, but teh obs_year variable still contains each of our three variables!

### `spread()`
```{r}
gap_normal <- gap_intermed %>% 
  spread(obs_type, obs_values) # spread out the "obs_type" column by the categories/levels and fill those columns with values from the column called obs_values
```

### All Together:
```{r}
gap_normal <- gap_wide %>% 
  gather(key = obstype_year, value = obs_values,
         -country, -continent) %>% 
  separate(obstype_year,
           into = c('obs_type', 'year'),
           sep = "_",
           convert = TRUE) %>% 
  spread(obs_type, obs_values)
```

### `unite()` (Going the Opposite Direction):
```{r}
gap_wide_new <- gapminder %>% 
  gather(key = obs_type, value = obs_values,
         -country, -continent, -year) %>%  # this give the gap_intermed dataframe
  unite(col = var_names, obs_type, year, sep = "_") %>% # this combines obs_type and year into a single column called var_names, concatenating the contents of the two rows with an underscore
  spread(key = var_names, value = obs_values) # spreads the data out based on the var_names column created above, fills with values from obs_values column
```












